Elephant Seal Age JAE Intraspecific Variation

### 2024-05-21 paper/paperv3.qmd
Trying segmented. No GLMM support tho :(

```{r}
#| label: setup
#| include: false  
library(ggthemes)
library(cowplot)
library(lme4)
library(lmerTest) # Loading this library will automatically append p-values to lmer models estimated using something called Satterthwaite's method
library(merTools)
library(tidyverse)
library(ggeffects)
library(patchwork)
source(here::here("R/format.R"))

sealdat <- read_csv(here::here("data/raw/128L pull 2023_12_05.csv"), 
                    show_col_types = FALSE) %>% 
  mutate(observed = if_else(observed == "B", "Breeder", "Non-breeder"), 
         observed_int = if_else(observed == "Breeder", 1, 0), 
         pup_survived = pupseeneveragain > 0) %>% 
  filter(age > 3, 
         year > 1987) %>% 
  mutate(across(c(animalID, year), factor),
         age10 = age / 10)

resight <- read.csv(here::here("data/raw/fullresights.csv")) %>% 
  dplyr::select(animalID, resightID, season, date, timeofyear, calyear, yday) %>% 
  mutate(animalID = as.factor(animalID))

@@

skip_frac <- sealdat %>% 
  distinct(animalID, year) %>% 
  mutate(year = as.numeric(as.character(year))) %>% 
  group_by(animalID) %>% 
  summarize(firstobs = min(year),
            lastobs = max(year)) %>% 
  mutate(n = lastobs - firstobs + 1) %>% 
  uncount(n) %>% 
  group_by(animalID) %>% 
  mutate(year = firstobs + seq(n()) - 1) %>% 
  ungroup() %>% 
  mutate(year = factor(year)) %>% 
  left_join(transmute(sealdat, animalID, year, skip = 0), by = c("animalID", "year")) %>% 
  mutate(skip = ifelse(is.na(skip), 1, 0)) %>% 
  group_by(year) %>% 
  summarize(skip_frac = sum(skip) / n())

resights_per_year %>% 
  mutate(year = factor(year)) %>% 
  left_join(skip_frac, by = "year") %>% 
  mutate(year = as.numeric(as.character(year))) %>% 
  ggplot(aes(median_resights, skip_frac)) +
  geom_point(aes(fill = year), shape = 21, color = "grey20", size = 2) +
  ggrepel::geom_label_repel(aes(label = year), max.overlaps = 8) +
  scale_y_continuous("Skip fraction", labels = scales::percent) +
  scale_fill_distiller(palette = "PuRd") +
  labs(x = "Median resights") +
  theme_classic()

@@

breeding_mod <- glmer(
  observed_int ~ poly(age10, 2, raw = TRUE) + (1 | animalID) + (1 | year),
  data = sealdat, 
  family = "binomial",
  control = glmerControl(optimizer="bobyqa")
)

@@

ranef_year <- ranef(breeding_mod)$year %>% 
  as_tibble(rownames = "year") %>% 
  arrange(`(Intercept)`) %>% 
  slice(1, floor(nrow(.) * c(0.25, 0.5, 0.75, 1))) %>% 
  expand_grid(age10 = unique(sealdat$age10)) %>% 
  mutate(predicted = predict(breeding_mod, newdata = ., re.form = ~ (1 | year), type = "response"),
         age = age10 * 10)

@@

@@

breed_nonbreed <- sealdat %>% 
  group_by(age) %>% 
  summarize(n_observed = sum(!is.na(observed)), 
            perc_breed = sum(observed == "Breeder") / n_observed, 
            se = (perc_breed * (1 - perc_breed) / n_observed)^0.5) %>% 
  filter(age < 22) %>%
  ggplot(aes(x = age, y = perc_breed)) + 
  geom_ribbon(aes(age, ymin = conf.low, ymax = conf.high),
              breed_pred,
              inherit.aes = FALSE,
              fill = "#829BD1",
              alpha = 0.25) +
  geom_line(aes(age, predicted, group = year),
            ranef_year,
            inherit.aes = FALSE,
            alpha = 0.2) +
  geom_text(aes(x = 22.1, predicted, label = year),
            filter(ranef_year, age == 22),
            inherit.aes = FALSE,
            hjust = 0,
            alpha = 0.5,
            size = 3) +
  geom_line(aes(age, predicted),
            breed_pred,
            inherit.aes = FALSE,
            color = "#829BD1",
            linewidth = 1) +
  geom_point() + 
  geom_errorbar(aes(ymin = perc_breed - se, ymax = perc_breed + se), width = 0) +
  scale_y_continuous(labels = scales::percent, limits = c(0.5, 1)) + 
  expand_limits(x = 23) +
  labs(x = "Maternal Age (Years)", y = "% Breeders (Mean ± SE)") + 
  theme_few() + 
  theme(axis.title = element_text(size = 13))

>>>>

```{r}
#| label: setup
#| include: false  
library(ggthemes)
library(cowplot)
library(lme4)
library(lmerTest) # Loading this library will automatically append p-values to lmer models estimated using something called Satterthwaite's method
library(merTools)
library(tidyverse)
library(ggeffects)
library(patchwork)
library(segmented)
source(here::here("R/format.R"))

sealdat <- read_csv(here::here("data/raw/128L pull 2023_12_05.csv"), 
                    show_col_types = FALSE) %>% 
  mutate(observed = if_else(observed == "B", "Breeder", "Non-breeder"), 
         observed_int = if_else(observed == "Breeder", 1, 0), 
         pup_survived = pupseeneveragain > 0) %>% 
  filter(age > 3, 
         year > 1987) %>% 
  mutate(animalID = factor(animalID),
         year_fct = factor(year),
         age10 = age / 10)

resight <- read.csv(here::here("data/raw/fullresights.csv")) %>% 
  select(animalID, resightID, season, date, timeofyear, calyear, yday) %>% 
  mutate(animalID = as.factor(animalID))

@@

skip_frac <- sealdat %>% 
  distinct(animalID, year) %>% 
  group_by(animalID) %>% 
  summarize(firstobs = min(year),
            lastobs = max(year)) %>% 
  mutate(n = lastobs - firstobs + 1) %>% 
  uncount(n) %>% 
  group_by(animalID) %>% 
  mutate(year = firstobs + seq(n()) - 1) %>% 
  ungroup() %>% 
  left_join(transmute(sealdat, animalID, year, skip = 0), 
            by = c("animalID", "year")) %>% 
  mutate(skip = ifelse(is.na(skip), 1, 0)) %>% 
  group_by(year) %>% 
  summarize(skip_frac = sum(skip) / n())

resights_per_year %>% 
  left_join(skip_frac, by = "year") %>% 
  ggplot(aes(median_resights, skip_frac)) +
  geom_point(aes(fill = year), shape = 21, color = "grey20", size = 2) +
  ggrepel::geom_label_repel(aes(label = year), max.overlaps = 8) +
  scale_y_continuous("Skip fraction", labels = scales::percent) +
  scale_fill_distiller(palette = "PuRd") +
  labs(x = "Median resights") +
  theme_classic()

@@

breeding_mod <- glmer(
  observed_int ~ poly(age10, 2, raw = TRUE) + (1 | animalID) + (1 | year_fct),
  data = sealdat, 
  family = "binomial",
  control = glmerControl(optimizer="bobyqa")
)

@@

ranef_year <- ranef(breeding_mod)$year_fct %>% 
  as_tibble(rownames = "year_fct") %>% 
  arrange(`(Intercept)`) %>% 
  slice(1, floor(nrow(.) * c(0.25, 0.5, 0.75, 1))) %>% 
  expand_grid(age10 = unique(sealdat$age10)) %>% 
  mutate(year_fct = factor(year_fct, levels = levels(sealdat$year_fct))) %>% 
  mutate(predicted = predict(breeding_mod, 
                             newdata = ., 
                             re.form = ~ (1 | year_fct), type = "response"),
         age = age10 * 10)

@@

Trying segmented analysis. Note: `segmented` does NOT support mixed effect GLMs!

```{r}
#| label: age_breeding_seg

breeding_mod2 <- glm(
  observed_int ~ age,
  data = sealdat, 
  family = "binomial"
)
breed_nseg <- lapply(
  list(15, c(9, 15)),
  \(.psi) {
    segmented(breeding_mod2, seg.Z = ~age, psi = .psi)
  }
)
AIC(breeding_mod2, breed_nseg[[1]], breed_nseg[[2]]) %>% 
  as_tibble() %>% 
  mutate(model = paste(1:3, "segments")) %>% 
  transmute(model, AIC, dAIC = AIC - min(AIC)) %>% 
  arrange(dAIC)

breeding_best <- breed_nseg[[1]]
breeding_pred <- tibble(age = seq(4, 22, length.out = 100)) %>% 
  predict(breeding_best, 
          newdata = ., 
          type = "response", 
          interval = "confidence") %>% 
  as_tibble() %>% 
  set_names(c("bred", "bred_lwr", "bred_upr")) %>% 
  mutate(age = seq(4, 22, length.out = 100))
breed_nonbreed <- sealdat %>% 
  group_by(age) %>% 
  summarize(n_observed = sum(!is.na(observed)), 
            perc_breed = sum(observed == "Breeder") / n_observed, 
            se = (perc_breed * (1 - perc_breed) / n_observed)^0.5) %>% 
  filter(age < 22)

ggplot(breeding_pred, aes(age)) +
  geom_ribbon(aes(ymin = bred_lwr, ymax = bred_upr),
              fill = "grey90") + 
  geom_line(aes(y = bred)) +
  geom_point(aes(y = perc_breed), breed_nonbreed) + 
  geom_errorbar(aes(ymin = perc_breed - se, ymax = perc_breed + se), 
                breed_nonbreed,
                width = 0) +
  theme_classic()

@@

breed_nonbreed <- sealdat %>% 
  group_by(age) %>% 
  summarize(n_observed = sum(!is.na(observed)), 
            perc_breed = sum(observed == "Breeder") / n_observed, 
            se = (perc_breed * (1 - perc_breed) / n_observed)^0.5) %>% 
  filter(age < 22) %>%
  ggplot(aes(x = age, y = perc_breed)) + 
  geom_ribbon(aes(age, ymin = conf.low, ymax = conf.high),
              breed_pred,
              inherit.aes = FALSE,
              fill = "#829BD1",
              alpha = 0.25) +
  geom_line(aes(age, predicted, group = year_fct),
            ranef_year,
            inherit.aes = FALSE,
            alpha = 0.2) +
  geom_text(aes(x = 22.1, predicted, label = year_fct),
            filter(ranef_year, age == 22),
            inherit.aes = FALSE,
            hjust = 0,
            alpha = 0.5,
            size = 3) +
  geom_line(aes(age, predicted),
            breed_pred,
            inherit.aes = FALSE,
            color = "#829BD1",
            linewidth = 1) +
  geom_point() + 
  geom_errorbar(aes(ymin = perc_breed - se, ymax = perc_breed + se), 
                width = 0) +
  scale_y_continuous(labels = scales::percent, limits = c(0.5, 1)) + 
  expand_limits(x = 23) +
  labs(x = "Maternal Age (Years)", y = "% Breeders (Mean ± SE)") + 
  theme_few() + 
  theme(axis.title = element_text(size = 13))
